### 2016-04-27 text/Chapter05.tex
Finish edits in Chapters 4 and 5

\newpage
~
\newpage
\paragraph{Properties.}
We now prove the type safety of of a context-aware programming language with implicit parameters.
To do this, we prove safety of the target functional language with specific extensions for implicit
parameters and we show that the translation from context-aware programming language with implicit
parameters produces well-typed programs in the target language.

@@

\newpage

@@

Indexed comonads are not restricted by this property of comonads. For example, given the indexed
product comonad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$, the shape of
the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
just implicit parameters $\cclrd{s}$.
Thus \emph{indexed comonads} are a generalization of \emph{comonads} that captures structures that
do not form a (non-indexed) comonad. However, indexing has been also discussed in the case of
\emph{monads.}

@@

\noindent
A \emph{family of comonads} is more restrictive than an \emph{indexed comonad}, because each of the data
types needs to form a comonad separately. For example, our indexed Maybe does not form a family of
comonads (again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given
a family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define
an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad,
we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.

\paragraph{Parameteric effect monads.}
Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our
indexed comonads) are closely related to our definition.  Although presented in a more general
categorical framework (and using monads), the model defines the \ident{unit} operation only on the
unit of a monoid and the \ident{bind} operation composes effect annotations using the provided monoidal
structure.

@@

\begin{itemize}
\item Semantically, coeffects capture different notions of computation. As demonstrated in
  Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
  computation, many of which cannot be captured by a monad (\eg~liveness or data-flow).

\item Syntactically, coeffect calculi use a richer algebraic structure with pointwise composition,
  sequential composition and context merging ($\cpar, \cseq$, and $\czip$) while most effect systems
  only use a single operation for sequential composition (used by monadic bind).

\item Syntactically, the second difference is in the lambda abstraction (\emph{abs}). In
  coeffect systems, the context demands of the body can be split between (or duplicated
  at) declaration site and call site, while monadic effect systems always defer all effects.
\end{itemize}

\noindent
Despite the differences, our implicit parameters example can be also represented by a monad.
Semantically, the \emph{reader} monad is equivalent to the \emph{product} comonad. Syntactically,
we use the $\cup$ operation for all three operations of the coeffect algebra. However, to enable
splitting of implicit parameter demands using the reader monad, we need to extend the
monad structure and change the translation of monadic lambda abstraction.

@@

\paragraph{Categorical relationship.}
Before looking at the necessary extensions, consider the two ways of modelling implicit
parameters. We assume that the function $\cclrd{r} \rightarrow \sigma$ is a lookup function
for reading implicit parameter values that is defined on a set $\cclrd{r}$. The two definitions
are:
%
\begin{equation*}
\begin{array}{lll}
 \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
 \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{reader monad})
\end{array}
\end{equation*}
%
The \emph{product comonad} simply pairs the value $\tau$ with the lookup function, while
the \emph{reader monad} is a function that, given a lookup function, produces a $\tau$ value.
As noted by Orchard \cite{comonads-vs-monads}, when used to model computation semantics, the
two representations are equivalent:
%
\begin{remark}
Computations modelled as $\ctyp{\cclrd{r}}{\tau_1}\rightarrow\tau_2$ using the product comonad
are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using the
reader monad via currying/uncurrying isomorphism.
\end{remark}
\begin{proof}
The isomorphism is demonstrated by the following equation:
\begin{equation*}
\begin{array}{lll}
 \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 &\narrow{=}&
 (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
 &\narrow{=}& \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) =
 \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \hspace{23em}\qedhere \\
\end{array}
\end{equation*}
\end{proof}

\noindent
This equivalence holds for monads and comonads (as well as \emph{indexed} monads
and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging.

@@

The operation cannot be implemented in a useful way for most standard monads, but the
reader monad is, indeed, an exception. It is not difficult to see how it can be implemented
when we expand the definitions of $\mtyp{\cclrd{r}}\tau$:
%
\begin{equation*}
\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~
(\tau_1\rightarrow(\cclrd{r \cup s} \rightarrow \sigma) \rightarrow \tau_2) \rightarrow
((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_1\rightarrow(\cclrd{s} \rightarrow \sigma) \rightarrow \tau_2)
\end{equation*}

@@

In the previous chapter, we defined \emph{type system} for flat coeffect calculi that uniformly
captures the shared structure of context-aware computations. In this chapter, we completed
the unification by providing \emph{semantics} for flat coeffect calculi and proving the
\emph{safety} of coeffect languages for dataflow and implicit parameters. The semantics shown
here also guides the implementation that is discussed later in Chapter~\ref{ch:impl}.

@@

We then used the comonadic semantics to define a \emph{comonadically-inspired translation}
that turns programs written in a domain-specific coeffect language into a functional target
language. This is akin to the Haskell do notation for monads. Finally, we extended the target
language with concrete implementation of comonadic operations for dataflow and implicit parameters
and we presented syntactic safety proof. In summary, the proof states that well-typed
context-aware programs written in a coeffect language \emph{do not go wrong} (when translated
to a simple functional language and evaluated).

>>>>

\paragraph{Properties.}
We now prove the type safety of of a context-aware programming language with implicit parameters.
To do this, we prove safety of the target functional language with specific extensions for implicit
parameters and we show that the translation from context-aware programming language with implicit
parameters produces well-typed programs in the target language.

@@

@@

Indexed comonads are not restricted by this property of comonads. For example, given the indexed
product comonad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$, the shape of
the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
just implicit parameters $\cclrd{s}$. Thus \emph{indexed comonads} are a generalization of
\emph{comonads} that captures structures that fail to form a comonad without indexing. In the
rest of the section, we look at work that discusses indexing in the context of \emph{monads.}

@@

\noindent
A \emph{family of comonads} is not as expressive as an \emph{indexed comonads}. Many indexed
comonads cannot be captured by a family of comonads. This is because each of the data
types needs to form a comonad separately. For example, our indexed Maybe does not form a family of
comonads (again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given
a family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define
an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad,
we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.

\paragraph{Parameteric effect monads.}
Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our
indexed comonads) are closely related to our definition.  Although presented in a more general
categorical framework (and using monads), the model (i) defines the \ident{unit} operation only on the
unit of a monoid and (ii) the \ident{bind} operation composes effect annotations using the provided
monoidal structure.

@@

\begin{itemize}
\item Semantically, coeffects capture different notions of computation. As demonstrated in
  Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
  computation, many of which cannot be captured by a monad (\eg~liveness or dataflow).
  In terms of program analysis \cite{other-optcomp}, monas capture forward dataflow analyses and
  comonads correspond to backward dataflow analyses.

\item Syntactically, coeffect calculi use a richer algebraic structure with pointwise composition,
  sequential composition and context merging ($\cpar, \cseq$, and $\czip$) while most effect systems
  only use a single operation for sequential composition (used by monadic bind). Effect systems
  may use a richer algebraic structure to support additional language constructs such as conditionals
  \cite{effects-nielson,effects-revisited}, but not for abstraction and application.

\item Syntactically, the second difference is in the lambda abstraction (\emph{abs}). In
  coeffect systems, the context demands of the body can be split between (or duplicated
  at) declaration site and call site, while lambda abstraction in monadic effect systems always
  defer all effects -- creating a function value has no effect.
\end{itemize}

\noindent
Despite the differences, our implicit paremeters resemble, in many ways, the \emph{reader} monad.
As discussed in Section~\ref{sec:semantics-related-monads}, the \emph{reader} monad is semantically
equivalent to the \emph{product} comonad when we consider just sequential composition. For a
language with lambda abstraction, we need a slight extension to the usual treatment of monads in
order to model implicit parameters using a monad.

@@

\paragraph{Categorical relationship.}
Before looking at the necessary extensions, consider the two ways of modelling implicit
parameters. We assume that the function $\cclrd{r} \rightarrow \ident{num}$ is a lookup function
for reading implicit parameter values that is defined on a set $\cclrd{r}$. The two definitions
are:
%
\begin{equation*}
\begin{array}{lll}
 \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
 \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{reader monad})
\end{array}
\end{equation*}
%
The \emph{product comonad} simply pairs the value $\tau$ with the lookup function, while
the \emph{reader monad} is a function that, given a lookup function, produces a $\tau$ value.
As noted by Orchard \cite{comonads-vs-monads}, when used to model computation semantics, the
two representations are equivalent:
%
\begin{remark}
Computations modelled as $\ctyp{\cclrd{r}}{\tau_1}\rightarrow\tau_2$ using the product comonad
are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using the
reader monad via currying/uncurrying isomorphism.
\end{remark}
\begin{proof}
The isomorphism is demonstrated by the following equation:
\begin{equation*}
\begin{array}{lll}
 \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 &\narrow{=}&
 (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
 &\narrow{=}& \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) =
 \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \hspace{23em}\qedhere \\
\end{array}
\end{equation*}
\end{proof}

\noindent
This equivalence shows an intriguing relationship between the \emph{product comonad} and
\emph{reader monad}, but it cannot be extended beyond that. In particular, comonads that model
dataflow computations or liveness do not have a corresponding monadic structure.
This equivalence holds for monads and comonads (as well as \emph{indexed} monads
and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging. This can be
supported in monadic computations by adding an aditional operation discussed next.

@@

The operation cannot be implemented in a useful way for most standard monads, but the
reader monad is, indeed, an exception. It is not difficult to see how it can be implemented
when we expand the definitions of $\mtyp{\cclrd{r}}\tau$:
%
\begin{equation*}
\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~
(\tau_1\rightarrow(\cclrd{r \cup s} \rightarrow \sigma) \rightarrow \tau_2) \rightarrow
((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_1\rightarrow(\cclrd{s} \rightarrow \sigma) \rightarrow \tau_2)
\end{equation*}
%
This suggests that the \emph{reader monad} is a special case among monads. Our work suggests that
passing read-only information to a computation is better captured by a product comonad, which also
matches the intuition -- read-only information is a \emph{contextual capability}.

@@

In the previous chapter, we defined a \emph{type system} for flat coeffect calculi that uniformly
captures the shared structure of context-aware computations. In this chapter, we completed
the unification by providing \emph{semantics} for flat coeffect calculi and proving the
\emph{safety} of coeffect languages for dataflow and implicit parameters. The semantics shown
here also guides the implementation that is discussed later in Chapter~\ref{ch:impl}.

@@

We then used the comonadic semantics to define a \emph{comonadically-inspired translation}
that turns programs written in a domain-specific coeffect language into a functional target
language. This is akin to the Haskell do-notation for monads. Finally, we extended the target
language with concrete implementations of comonadic operations for dataflow and implicit parameters
and we presented a syntactic safety proof. In summary, the proof states that well-typed
context-aware programs written in a coeffect language \emph{do not go wrong} (when translated
to a simple functional language and evaluated).
